Comparison of Terrain Following and Cut Cell Grids using a Non-Hydrostatic Model

### 2015-08-20 src/tf-cutcell-comparison/tf-cutcell-comparison.tex
minor corrections and revisions from reviewer

\bibpunct{(}{)}{;}{a}{}{,}

@@

\abstract{Terrain following coordinates are widely used in operational models but the cut cell method has been proposed as an alternative that can more accurately represent atmospheric dynamics over steep orography.  Because the type of grid is usually chosen during model implementation, it is typically necessary to use different models in order to compare the accuracy of different grids.  In contrast, here a single C-grid finite volume model is used to enable a like-for-like comparison of terrain following and cut cell grids.
A series of standard two-dimensional tests using idealised terrain are performed: tracer advection in a prescribed horizontal velocity field, a test starting from stably stratified initial conditions, and orographically induced gravity waves described by nonhydrostatic dynamics.  In addition, a new tracer advection test is formulated having a velocity field that is everywhere tangential to the terrain following coordinate surfaces.  This new test presents a challenge to cut cells in situations with a large Froude number and large Scorer parameter.
The results of the advection tests demonstrate that tracer accuracy depends upon alignment of the flow with the grid.  As expected, the cut cell grid maintains greater accuracy in the test of stationary, stratified flow.  In the gravity waves test, results on all grids are in good agreement with existing results from the literature.  However, it is found that the Lorenz computational mode is excited as a vertical zig--zag in potential temperature on the cut cell grid.}

@@

\section{Introduction}
Representing orography accurately in numerical weather prediction systems is necessary to model downslope winds and local precipitation.  Orography also exerts strong non-local influences: from the latent heat release due to convection, by directly forcing planetary waves, and by the atmospheric response to form drag and gravity wave drag.  There are two main approaches to represent orography on a grid: terrain following layers and cut cells.  Both methods align cells in vertical columns.  Because most models are designed for a particular type of grid, existing studies of cut cell solutions have compared results with terrain following grid solutions implemented within different models, for example \citet{good2014}.  Instead, this study uses a single model to enable a like-for-like comparison between solutions using terrain following and cut cell grids.

@@

Second, numerical errors can also be reduced by improving the accuracy in calculating the horizontal pressure gradient itself.  TF layers are usually implemented using a coordinate transformation onto a rectangular computational domain, which introduces metric terms into the equations of motion.  The techniques proposed by \citet{klemp2011} and \cite{zaengl2012} both involve the calculation of the horizontal pressure gradient in the physical domain.  This gave them the flexibility to design more accurate horizontal pressure gradient discretizations using more appropriate stencils.

Despite their associated numerical errors, TF layers are in widespread operational use \citep{steppeler2003}.  They are attractive because their rectangular structure is simple to process by computer and link with parameterisations, boundary layer resolution can be increased with variable spacing of vertical layers \citep{schaer2002}, and cell sizes remain almost constant \citep{jebens2011}.

@@

The cut cell method can create some very small cells which reduce computational efficiency \citep{klein2009}, and several approaches have been tried to alleviate the problem \citep{steppeler2002,yamazaki-satomura2010,jebens2011}.

@@

In an experiment of orographically induced gravity waves, \citet{gallus-klemp2000} found that horizontal flow along the lee slope was artificially weak.  \citet{mesinger2012} offer an explanation for this behaviour: air flowing along the lee slope cannot travel diagonally downwards but must first travel horizontally, then vertically downward.  However, lee slope winds are weakened because some of the air continues to be transported horizontally aloft.

@@

\citet{galchen-somerville1975} proposed a basic terrain following (BTF) coordinate defined as 
\begin{equation}
	z = \left( H - h \right) \left( z^\star / H \right) + h \label{eqn:btf}
\end{equation}
where, in two dimensions, \(z(x, z^\star)\) is the physical height of the coordinate surface at level \(z^\star\), \(H\) is the height of the domain, and \(h(x)\) is the height of the terrain surface.  In this formulation $z^\star/H$ ranges from 0 to 1.  Using this coordinate, the terrain's influence decays linearly with height but disappears only at the top of the domain.  An example is shown in figure~\ref{fig:grid}a.

The sigma coordinate transform of \citet{phillips1957} is equivalent to the BTF coordinate transform since they both decay linearly.  However, because \(\sigma = p/p_s\) decays with pressure rather than height, sigma coordinates also vary with surface pressure, \(p_s(x, t)\).

The hybrid terrain following (HTF) coordinates of \citet{simmons-burridge1981} improve upon BTF coordinates by using a decay function that allows the influence of the terrain to be removed at a specified height, producing model layers aloft that have constant pressure.

@@

Most implementations of terrain following layers use a coordinate system that makes the domain rectangular, but introduces metric terms into the equations of motion.  Instead, the model employed in this study uses Cartesian coordinates and unstructured grids.  By doing so, results from the same model can be compared between terrain following and cut cell grids without modifying the equation set or discretisation.

@@

\section{Model}
\label{sec:model}

We use the finite volume model from \citet{weller-shahrokhi2014} which details a C-grid discretisation of the fully-compressible Euler equations, given by
\begin{subequations}
\begin{align}
	\text{Momentum} &\ &\  	\frac{\partial \rho \bm{u}}{\partial t} + \nabla \cdot \rho \bm{uu} &= \rho \bm{g} - c_p \rho \theta \nabla \Pi \label{eq:momentum} \\
	\text{Continuity} &\ &\	\frac{\partial \rho}{\partial t} + \nabla \cdot \rho \bm{u} &= 0 \\
	\text{Thermodynamic equation} &\ &\ \frac{\partial \rho \theta}{\partial t} + \nabla \cdot \rho \bm{u} \theta &= 0 \\
	\text{Ideal gas law} &\ &\ \Pi^{(1 - \kappa)/\kappa} &= \frac{R \rho \theta}{p_0}
\end{align}
\end{subequations}
where \(\rho\) is the density, \(\bm{u}\) is the velocity field, \(\bm{g}\) is the gravitational acceleration, \(c_p\) is the heat capacity at constant pressure, \(\theta = T \left(p_0/p\right)^\kappa\) is the potential temperature, \(T\) is the tempereature, \(p\) is the pressure, \(p_0\) is a reference pressure, \(\Pi = \left(p / p_0 \right)^\kappa\) is the Exner function of pressure, and \(\kappa = R/c_p\) is the gas constant to heat capacity ratio.

The model uses a Lorenz C-grid staggering of thermodynamic variables such that $\theta$, $\rho$ and $\Pi$ are stored at cell centres and the covariant component of velocity at cell faces.

A minor change has been made to the curl-free, non-orthogonal formulation as described in \citet{weller-shahrokhi2014} to improve stability on non-orthogonal grids.  \citet{weller-shahrokhi2014} use a Hodge dual operator, \(H\), to map from co-variant to contra-variant velocity components:
\begin{align}
	U = HV
\end{align}
where \(U = \rho\bm{u}\cdot\bm{S}_f\) and \(V=\rho\bm{u}\cdot\bm{d}_f\).  \(\bm{S}_f\) is the surface normal vector for a face, \(f\), which has a magnitude of the surface area, and \(\bm{d}_f\) is the vector between neighbouring cell centres.

@@

For tracer advection tests, the advection equation in flux form, given by
\begin{align}
\partial \phi / \partial t + \nabla \cdot \left( \bm{u} \phi \right) = 0
\end{align}
is solved using the upwind-biased multidimensional cubic scheme from \citet{weller-shahrokhi2014} which is non-monotonic and not flux corrected.
The time derivative is solved using a three-stage, second order Runge-Kutta scheme:
\begin{subequations}
\begin{align}
	\varphi^\star &= \varphi^{(n)} + \Delta t f(\phi^{(n)}) \\
	\varphi^{\star\star} &= \varphi^{(n)} + \frac{\Delta t}{2} \left( f(\phi^{(n)}) + f(\phi^\star) \right) \\
	\varphi^{(n+1)} &= \varphi^{(n)} + \frac{\Delta t}{2} \left( f(\phi^{(n)}) + f(\phi^{\star\star}) \right)
\end{align}
\end{subequations}
where \(f(\phi^{(n)}) = - \nabla \cdot (\bm{u} \varphi^{(n)})\) at time level \(n\).

@@

The terrain is wave-shaped, specified by the surface height, \(h\), such that
\begin{subequations}
\begin{align}
   h(x) &= h^\star \cos^2 ( \alpha x )
%
\intertext{where}
%
   h^\star(x) &= \left\{ \begin{array}{l l}
       h_0 \cos^2 ( \beta x ) & \quad \text{if $| x | < a$} \\
	0 & \quad \text{otherwise}
    \end{array} \right.
\end{align}
\end{subequations}
where $a = \SI{25}{\kilo\meter}$ is the mountain envelope half-width, $h_0 = \SI{3}{\kilo\meter}$ is the maximum mountain height, $\lambda = \SI{8}{\kilo\meter}$ is the wavelength, \(\alpha = \pi / \lambda\) and \(\beta = \pi / 2a\).  On the SLEVE grid, the large-scale component $h_1$ is given by \(h_1(x) = h^\star(x) / 2\)
and $s_1 = \SI{15}{\kilo\meter}$ is the large scale height, and $s_2 = \SI{2.5}{\kilo\meter}$ is the small scale height.  The optimisation of SLEVE by \citet{leuenberger2010} is not used, so the exponent $n = 1$.

The wind is entirely horizontal and is prescribed as
\begin{align}
	u(z) = u_0 \left\{ \begin{array}{l l}
		1 & \quad \text{if $z \geq z_2$} \\
		\sin^2 \left( \frac{\pi}{2} \frac{z - z_1}{z_2 - z_1} \right) & \quad \text{if $z_1 < z < z_2$} \\
		0 & \quad \text{otherwise}
	\end{array} \right.	
\end{align}
where $u_0 = \SI{10}{\meter\per\second}$, $z_1 = \SI{4}{\kilo\meter}$ and $z_2 = \SI{5}{\kilo\meter}$.
This results in a constant wind aloft, and zero flow at \SI{4}{\kilo\meter} and below.

To ensure that the discrete velocity field is non-divergent, velocities are prescribed at cell faces by differencing the streamfunction, \(\Psi\), along the edges from \(\Psi\) stored at cell vertices.  Since \(u = - \partial \Psi / \partial z\), the streamfunction is found by vertical integration of the velocity profile:
\begin{align}
	\Psi(z) &= \frac{u_0}{2} \left\{ \begin{array}{l l}
		\left( 2z - z_1 - z_2 \right) & \enskip \text{if $z > z_2$} \\
		z - z_1 - \frac{z_2 - z_1}{\pi} \sin \left(\pi \frac{z - z_1}{z_2-z_1}\right) & \enskip \text{if $z_1 < z \leq z_2$} \\
		0 & \enskip \text{if $z \leq z_1$}
	\end{array} \right.
\end{align}

A tracer, $\varphi$, is positioned upstream above the height of the terrain.  It has the shape
\begin{align}
	\varphi(x, z) &= \varphi_0 \left\{ \begin{array}{l l}
		\cos^2 \left( \frac{\pi r}{2} \right) & \quad \text{if $r \leq 1$} \\
		0 & \quad \text{otherwise}
	\end{array} \right.
%
\intertext{having radius, $r$, given by}
%
	r &= \sqrt{
		\left( \frac{x - x_0}{A_x} \right)^2 + 
		\left( \frac{z - z_0}{A_z} \right)^2
	}
\end{align}
where $A_x = \SI{25}{\kilo\meter}$, $A_z = \SI{3}{\kilo\meter}$ are the horizontal and vertical half-widths respectively, and $\varphi_0 = 1$ is the maximum magnitude of the anomaly.  At $t = \SI{0}{\second}$, the anomaly is centred at $(x_0, z_0) = (\SI{-50}{\kilo\meter}, \SI{9}{\kilo\meter})$ so that the anomaly is upwind of the mountain and well above the maximum terrain height of \SI{3}{\kilo\meter}.  Analytic solutions can be found by setting the anomaly centre such that $x_0 = ut$.

@@

The test was executed on the BTF, SLEVE and cut cell grids, and on a regular grid with flat terrain using a centred linear scheme and the upwind-biased cubic scheme.   Results were also obtained on BTF and SLEVE grids with the fourth order scheme from \citet{schaer2002} using a modified version of their Fortran code.  The original version of the code interpolated the geometric height at cell vertices from values at adjacent half levels in order to calculate the streamfunction, \(\Psi(z)\).  The modified version used here directly calculates the height at vertices to enable comparisons with the finite volume model solutions.

@@

By \(t = \SI{10000}{\second}\), the tracer suffers from distortion on the BTF grid using the centred linear scheme and some artefacts remain about the mountain peak.  The tracer has spread vertically due to increased numerical errors when the tracer is transported between layers.  Distortions are reduced by using the fourth order scheme from \citet{schaer2002} (figure~\ref{fig:advection-tracer}c), but the computational mode is seen as a grid-scale oscillation that travels in the opposite direction to the wind (figure~\ref{fig:advection-tracer}e).  The results from the centred linear and fourth order schemes are slightly worse than the respective results from \citet{schaer2002} (their figure 6a and figure 8) because of the difference in the discretisation of $\Psi$.

@@

Minimum and maximum tracer values and \(\ell^2\) error norms on the BTF, SLEVE, cut cell and regular grids are summarised in table~\ref{tab:advection}, where the \(\ell^2\) error norm is defined as 
\begin{align}
	\ell^2 &= \sqrt{\frac{\sum_c \left( \varphi - \varphi_{T} \right)^2 \mathcal{V}_c}{\sum_c \left( \varphi_T^2 \mathcal{V}_c \right)}} \label{eqn:l2-error}
\end{align}
where $\varphi$ is the numerical tracer value, $\varphi_T$ is the analytic value and $\mathcal{V}_c$ is the cell volume.

The results of the cubic upwind-biased scheme on TF and regular grids are comparable with those for the fourth-order centred scheme from \citet{schaer2002}.  Error is largest on the BTF grid with \(\ell^2 = \num{0.00791}\) but significant reduced on the SLEVE grid with \(\ell^2 = \num{0.00108}\).  The error is approximately halved by changing from the SLEVE grid to the cut cell grid.
Tracer minima and maxima for the centred linear and fourth order schemes are lower than those presented by \citet{schaer2002} because no interpolation is used to calculate the streamfunction.

The centred linear scheme is sensitive to changes in the terrain profile: when the domain width is reduced to \SI{300}{\kilo\meter}, the $\ell^2$ error on the BTF grid rises from \num{0.0210} to \num{0.0318}.  The upwind-biased cubic scheme is less sensitive to the same change, with the $\ell^2$ error decreasing from \num{0.00791} to \num{0.00767}.

@@

$\ell^2$ errors and tracer extrema for this test are compared with the horizontal advection results in table~\ref{tab:advection}.  In the terrain following velocity field, tracer accuracy is greatest on the BTF grid.  Using the cubic upwind-biased scheme, errors are about ten times larger on the SLEVE and cut cell grids compared to the BTF grid.

Using the centred linear scheme on the cut cell grid, numerical instability is generated by flow through small cells: when interpolating \(\phi\) onto faces, flux from a large cell into a small cell means that the downwind cell is given greater weighting.  The maximum courant number is less than one in all of the advection tests, hence it is not the source of this instability.

@@

The test setup follows the specification by \cite{klemp2011}.  The grid resolution is \(\Delta x = \Delta z^\star = \SI{500}{\meter}\).  All boundary conditions are no normal flow.

@@

Following \citet{melvin2010}, the domain is \SI{300}{\kilo\meter} wide and \SI{30}{\kilo\meter} high.  The mountain profile has the same form as equation~\ref{eqn:resting:mountain}.  Tests are performed with mountain heights of $h_0 = \SI{250}{\meter}$ and \(h_0 = \SI{500}{\meter}\).  As in the resting atmosphere test, $a = \SI{5}{\kilo\meter}$ is the mountain half-width and $\lambda = \SI{4}{\kilo\meter}$ is the wavelength.  On the optimised SLEVE grid, $s_1 = \SI{5}{\kilo\meter}$ is the large scale height, $s_2 = \SI{2}{\kilo\meter}$ is the small scale height and the optimal exponent value $n = 1.35$ is used.

>>>>

\bibpunct{(}{)}{;}{a}{}{,}
\newcommand{\TODO}[1]{\textcolor{purple}{TODO: \emph{#1}}}

@@

\abstract{Terrain following coordinates are widely used in operational models but the cut cell method has been proposed as an alternative that can more accurately represent atmospheric dynamics over steep orography.  Because the type of grid is usually chosen during model implementation, it is typically necessary to use different models in order to compare the accuracy of different grids.  In contrast, here a single C-grid finite volume model is used to enable a like-for-like comparison of terrain following and cut cell grids.
A series of standard two-dimensional tests using idealised terrain are performed: tracer advection in a prescribed horizontal velocity field, a test starting from stably stratified initial conditions, and orographically induced gravity waves described by nonhydrostatic dynamics.  In addition, a new tracer advection test is formulated having a velocity field that is everywhere tangential to the terrain following coordinate surfaces.  This new test presents a challenge to cut cells in situations with a large Froude number and large Scorer parameter.  \TODO{discuss Froude number/Scorer parameter in body.}
The results of the advection tests demonstrate that tracer accuracy depends upon alignment of the flow with the grid.  As expected, the cut cell grid maintains greater accuracy in the test of stationary, stratified flow.  In the gravity waves test, results on all grids are in good agreement with existing results from the literature.  However, it is found that the Lorenz computational mode is excited as a vertical zig--zag in potential temperature on the cut cell grid.}

@@

\section{Introduction}
Representing orography accurately in numerical weather prediction systems is necessary to model downslope winds and local precipitation.  Orography also exerts strong non-local influences: from the latent heat release due to convection, by directly forcing planetary waves, and by the atmospheric response to form drag and gravity wave drag.  There are two main approaches to representing orography on a grid: terrain following layers and cut cells.  \TODO{another way of treating orography are 'immersed boundary methods' or 'embedded boundary methods'}  Both \TODO{all?} methods align cells in vertical columns.  Because most models are designed for a particular type of grid, existing studies of cut cell solutions have compared results with terrain following grid solutions implemented within different models, for example \citet{good2014}.  Instead, this study uses a single model to enable a like-for-like comparison between solutions using terrain following and cut cell grids.

@@

Second, numerical errors can also be reduced by improving the accuracy in calculating the horizontal pressure gradient itself.  TF layers are usually implemented using a coordinate transformation onto a rectangular computational domain, which introduces metric terms into the equations of motion.  The techniques proposed by \citet{klemp2011} and \cite{zaengl2012} both involve the calculation of the horizontal pressure gradient in the physical domain \TODO{interpolate to z-levels in order to compute the horizontal pressure gradient}.  This gave them the flexibility to design more accurate horizontal pressure gradient discretizations using more appropriate stencils.

Despite their associated numerical errors, TF layers are in widespread operational use \citep{steppeler2003}.  They are attractive because their rectangular structure is simple to process by computer and link with parameterisations, boundary layer resolution can be increased with variable spacing of vertical layers \citep{schaer2002}, and cell sizes remain almost constant \citep{jebens2011}. \TODO{clarify; constant cell sizes but variable spacing of vertical layers}

@@

The cut cell method can create some very small cells which reduce computational efficiency \citep{klein2009}, and several approaches have been tried to alleviate the problem.  \citet{yamazaki-satomura2010} combine small cells with horizontally or vertically adjacent cells.  \citet{steppeler2002} employ a thin-wall appoximation to increase the computational volume of small cells without altering the terrain.  \citet{jebens2011} avoid the timestep restriction associated with explicit schemes by using an implicit method for cut cells and a semi-explicit method elsewhere.

@@

In an experiment of orographically induced gravity waves, \citet{gallus-klemp2000} found that horizontal flow along the lee slope was artificially weak in the Eta model.  \citet{mesinger2012} offer an explanation for this behaviour: air flowing along the lee slope cannot travel diagonally downwards but must first travel horizontally, then vertically downward.  However, lee slope winds are weakened because some of the air continues to be transported horizontally aloft.

@@

\citet{galchen-somerville1975} proposed a basic terrain following (BTF) coordinate defined as 
\begin{equation}
	z = \left( H - h \right) \left( z^\star / H \right) + h \label{eqn:btf}
\end{equation}
where, in two dimensions, \(z(x, z^\star)\) is the physical height of the Cartesian coordinate surface at the model level with transformed height \(z^\star\), \(H\) is the height of the domain, and \(h(x)\) is the height of the terrain surface.  In this formulation $z$ varies between $h$ and $H$ and $z^\star$ ranges from 0 to $H$.  Using this coordinate, the terrain's influence decays linearly with height but disappears only at the top of the domain.  An example is shown in figure~\ref{fig:grid}a.

@@

Most implementations of terrain following layers use a coordinate system that makes the domain rectangular, but introduces metric terms into the equations of motion.  Instead, the model employed in this study uses Cartesian coordinates and unstructured grids \TODO{they're very structured!}.  By doing so, results from the same model can be compared between terrain following and cut cell grids without modifying the equation set or discretisation.

@@

\section{Model}
\label{sec:model}
\TODO{Use model in 2D x-z slice configuration without Coriolis forces}
\TODO{Add info about the different numerical options that are assessed in combination with the orography treatment}

We use the finite volume model from \citet{weller-shahrokhi2014} which details a discretisation of the fully-compressible Euler equations, given by
\begin{subequations}
\begin{align}
	\text{Momentum} &\ &\  	\frac{\partial \rho \bm{u}}{\partial t} + \nabla \cdot \rho \bm{uu} &= \rho \bm{g} - c_p \rho \theta \nabla \Pi \label{eq:momentum} \\
	\text{Continuity} &\ &\	\frac{\partial \rho}{\partial t} + \nabla \cdot \rho \bm{u} &= 0 \\
	\text{Thermodynamic equation} &\ &\ \frac{\partial \rho \theta}{\partial t} + \nabla \cdot \rho \bm{u} \theta &= 0 \\
	\text{Ideal gas law} &\ &\ \Pi^{(1 - \kappa)/\kappa} &= \frac{R \rho \theta}{p_0}
\end{align}
\end{subequations}
where \(\rho\) is the density, \(\bm{u}\) is the velocity field, \(\bm{g}\) is the gravitational acceleration, \(c_p\) is the heat capacity at constant pressure, \(\theta = T \left(p_0/p\right)^\kappa\) is the potential temperature, \(T\) is the temperature, \(p\) is the pressure, \(p_0\) is a reference pressure, \(\Pi = \left(p / p_0 \right)^\kappa\) is the Exner function of pressure, and \(\kappa = R/c_p\) is the gas constant to heat capacity ratio.

The model uses C-grid staggering in the horizontal and Lorenz staggering in the vertical such that $\theta$, $\rho$ and $\Pi$ are stored at cell centres and the covariant component of velocity at cell faces.

\TODO{move this to appendix and clarify}
A minor change has been made to the curl-free, non-orthogonal formulation as described in \citet{weller-shahrokhi2014} to improve stability on non-orthogonal grids.  \citet{weller-shahrokhi2014} use a Hodge dual operator, \(H\), to map from co-variant to contra-variant velocity components:
\begin{align}
	U = HV
\end{align}
where \(U = \rho\bm{u}\cdot\bm{S}_f\) and \(V=\rho\bm{u}\cdot\bm{d}_f\).  \(\bm{S}_f\) is the surface normal vector for a face, \(f\), which has a magnitude of the surface area, and \(\bm{d}_f\) is the vector between neighbouring cell centres.

@@

For tracer advection tests, the advection equation is solved in flux form, given by
\begin{align}
\partial \phi / \partial t + \nabla \cdot \left( \bm{u} \phi \right) = 0
\end{align}
where the tracer density, $\phi$, is interpolated onto cell faces using the upwind-biased multidimensional cubic scheme from \citet{weller-shahrokhi2014} which is non-monotonic and not flux corrected.
The time derivative is solved using a three-stage, second order Runge-Kutta scheme:
\begin{subequations}
\begin{align}
	\phi^\star &= \phi^{(n)} + \Delta t f(\phi^{(n)}) \\
	\phi^{\star\star} &= \phi^{(n)} + \frac{\Delta t}{2} \left( f(\phi^{(n)}) + f(\phi^\star) \right) \\
	\phi^{(n+1)} &= \phi^{(n)} + \frac{\Delta t}{2} \left( f(\phi^{(n)}) + f(\phi^{\star\star}) \right)
\end{align}
\end{subequations}
where \(f(\phi^{(n)}) = - \nabla \cdot (\bm{u} \phi^{(n)})\) at time level \(n\).

@@

The terrain is wave-shaped, specified by the surface height, \(h\), such that
\begin{subequations}
\begin{align}
   h(x) &= h^\star \cos^2 ( \alpha x )
%
\intertext{where}
%
   h^\star(x) &= \left\{ \begin{array}{l l}
       h_0 \cos^2 ( \beta x ) & \quad \text{if $| x | < a$} \\
	0 & \quad \text{otherwise}
    \end{array} \right.
\end{align}
\end{subequations}
where $a = \SI{25}{\kilo\meter}$ is the mountain envelope half-width, $h_0 = \SI{3}{\kilo\meter}$ is the maximum mountain height, $\lambda = \SI{8}{\kilo\meter}$ is the wavelength, \(\alpha = \pi / \lambda\) and \(\beta = \pi / (2a)\).  On the SLEVE grid, the large-scale component $h_1$ is given by \(h_1(x) = h^\star(x) / 2\)
and $s_1 = \SI{15}{\kilo\meter}$ is the large scale height, and $s_2 = \SI{2.5}{\kilo\meter}$ is the small scale height.  The optimisation of SLEVE by \citet{leuenberger2010} is not used, so the exponent $n = 1$.

The wind is entirely horizontal and is prescribed as
\begin{align}
	u(z) = u_0 \left\{ \begin{array}{l l}
		1 & \quad \text{if $z \geq z_2$} \\
		\sin^2 \left( \frac{\pi}{2} \frac{z - z_1}{z_2 - z_1} \right) & \quad \text{if $z_1 < z < z_2$} \\
		0 & \quad \text{otherwise}
	\end{array} \right.	
\end{align}
where $u_0 = \SI{10}{\meter\per\second}$, $z_1 = \SI{4}{\kilo\meter}$ and $z_2 = \SI{5}{\kilo\meter}$.
This results in a constant wind above $z_2$, and zero flow at \SI{4}{\kilo\meter} and below.

\TODO{double check signs of all streamfunctions (in horizontal and TF advection test cases)}
To ensure that the discrete velocity field is non-divergent, velocities are prescribed at cell faces by differencing the streamfunction, \(\Psi\), along the edges from \(\Psi\) stored at cell vertices.  Since \(u = \partial \Psi / \partial z\), the streamfunction is found by vertical integration of the velocity profile:
\begin{align}
	\Psi(z) &= \frac{u_0}{2} \left\{ \begin{array}{l l}
		\left( 2z - z_1 - z_2 \right) & \enskip \text{if $z > z_2$} \\
		z - z_1 - \frac{z_2 - z_1}{\pi} \sin \left(\pi \frac{z - z_1}{z_2-z_1}\right) & \enskip \text{if $z_1 < z \leq z_2$} \\
		0 & \enskip \text{if $z \leq z_1$}
	\end{array} \right.
\end{align}

A tracer with density $\phi$ is positioned upstream above the height of the terrain.  It has the shape
\begin{align}
	\phi(x, z) &= \phi_0 \left\{ \begin{array}{l l}
		\cos^2 \left( \frac{\pi r}{2} \right) & \quad \text{if $r \leq 1$} \\
		0 & \quad \text{otherwise}
	\end{array} \right.
%
\intertext{having radius, $r$, given by}
%
	r &= \sqrt{
		\left( \frac{x - x_0}{A_x} \right)^2 + 
		\left( \frac{z - z_0}{A_z} \right)^2
	}
\end{align}
where $A_x = \SI{25}{\kilo\meter}$, $A_z = \SI{3}{\kilo\meter}$ are the horizontal and vertical half-widths respectively, and $\phi_0 = 1$ is the maximum density of the tracer.  At $t = \SI{0}{\second}$, the tracer is centred at $(x_0, z_0) = (\SI{-50}{\kilo\meter}, \SI{9}{\kilo\meter})$ so that the tracer is upwind of the mountain and well above the maximum terrain height of \SI{3}{\kilo\meter}.  Analytic solutions can be found by setting the tracer centre such that $x_0 = ut$.

@@

The test was executed on the BTF, SLEVE and cut cell grids, and on a regular grid with flat terrain using a centred linear scheme and the upwind-biased cubic scheme.   Results were also obtained on BTF and SLEVE grids with the fourth order scheme from \citet{schaer2002} using a modified version of their Fortran code.  The original version of the code interpolated the geometric height at cell vertices from values at adjacent half levels in order to calculate the streamfunction, \(\Psi(z)\).  The modified version used here directly calculates the height at vertices to enable comparisons with the finite volume model solutions.  \TODO{move this detail into section 3 and add more info and references}

@@

By \(t = \SI{10000}{\second}\), the tracer suffers from distortion on the BTF grid using the centred linear scheme and some artifacts remain about the mountain peak.  The tracer has spread vertically due to increased numerical errors when the tracer is transported between layers.  Distortions are reduced by using the fourth order scheme from \citet{schaer2002} (figure~\ref{fig:advection-tracer}c), but the computational mode is seen as a grid-scale oscillation that travels in the opposite direction to the wind (figure~\ref{fig:advection-tracer}e).  The results from the centred linear and fourth order schemes are slightly worse than the respective results from \citet{schaer2002} (their figure 6a and figure 8) because of the difference in the discretisation of $\Psi$.

@@

Minimum and maximum tracer values and \(\ell_2\) error norms on the BTF, SLEVE, cut cell and regular grids are summarised in table~\ref{tab:advection}, where the \(\ell_2\) error norm is defined as 
\begin{align}
	\ell_2 &= \sqrt{\frac{\sum_c \left( \phi - \phi_{T} \right)^2 \mathcal{V}_c}{\sum_c \left( \phi_T^2 \mathcal{V}_c \right)}} \label{eqn:l2-error}
\end{align}
where $\phi$ is the numerical tracer value, $\phi_T$ is the analytic value and $\mathcal{V}_c$ is the cell volume.

The results of the cubic upwind-biased scheme on TF and regular grids are comparable with those for the fourth-order centred scheme from \citet{schaer2002}.  Error is largest on the BTF grid with \(\ell_2 = \num{0.107}\) but is significantly reduced on the SLEVE grid with \(\ell_2 = \num{0.0146}\).  The error is approximately halved by changing from the SLEVE grid to the cut cell grid.
Tracer minima and maxima for the centred linear and fourth order schemes are lower than those presented by \citet{schaer2002} because no interpolation is used to calculate the streamfunction.

The centred linear scheme is sensitive to changes in the terrain profile: when the domain width is reduced to \SI{300}{\kilo\meter}, the $\ell_2$ error on the BTF grid rises from \num{0.284} to \num{0.432}.  The upwind-biased cubic scheme is less sensitive to the same change, with the $\ell_2$ error decreasing from \num{0.107} to \num{0.104}.

@@

$\ell_2$ errors and tracer extrema for this test are compared with the horizontal advection results in table~\ref{tab:advection}.  In the terrain following velocity field, tracer accuracy is greatest on the BTF grid.  Using the cubic upwind-biased scheme, errors are about ten times larger on the SLEVE and cut cell grids compared to the BTF grid.

Using the centred linear scheme on the cut cell grid, numerical instability is generated by flow through small cells: when interpolating \(\phi\) onto faces, flux from a large cell into a small cell means that the downwind cell is given greater weighting.  The maximum Courant number is less than one in all of the advection tests, hence it is not the source of this instability.

@@

The test setup follows the specification by \cite{klemp2011}.  \TODO{how large is the domain? do I say this elsewhere? if not, say it here} The grid resolution is \(\Delta x = \Delta z^\star = \SI{500}{\meter}\).  All boundary conditions are no normal flow.

@@

Following \citet{melvin2010}, the domain is \SI{300}{\kilo\meter} wide and \SI{30}{\kilo\meter} high.  \TODO{state grid spacing} The mountain profile has the same form as equation~\ref{eqn:resting:mountain}.  Tests are performed with mountain heights of $h_0 = \SI{250}{\meter}$ and \(h_0 = \SI{500}{\meter}\).  As in the resting atmosphere test, $a = \SI{5}{\kilo\meter}$ is the mountain half-width and $\lambda = \SI{4}{\kilo\meter}$ is the wavelength.  On the optimised SLEVE grid, $s_1 = \SI{5}{\kilo\meter}$ is the large scale height, $s_2 = \SI{2}{\kilo\meter}$ is the small scale height and the optimal exponent value $n = 1.35$ is used.
